Visualizing Independence Using Extended Association Plots

### 2004-08-24 trunk/vcd/paper/JCGS.Rnw
Added several 3-way and 4-way tables from Andersen (1991)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@298 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
*** empty log message ***
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@301 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% Hyperref
\hypersetup{%
  pdftitle = {Visualizing (Conditional) Independence Using Extended Association and Mosaic Plots},
  pdfsubject = {Visualizing (Conditional) Independence},
  pdfkeywords = {contingency tables, association plots, mosaic plots, conditional
    inference, permutation tests},
  pdfauthor = {Achim Zeileis, David Meyer, Kurt Hornik},
  %% change colorlinks to false for pretty printing
  colorlinks = {true},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

@@

\begin{center}
\textbf{\LARGE Visualizing (Conditional) Independence Using Extended Association and Mosaic Plots}

@@

data(Bundesliga)
goals <- xtabs(~ AwayGoals + HomeGoals, data = Bundesliga, subset = Year == 1995)

data(HairEyeColor)
haireye <- xtabs(Freq ~ Hair + Eye, data = as.data.frame(HairEyeColor), subset = Sex == "Female")

data(UCBAdmissions)
ucb <- UCBAdmissions
dimnames(ucb)[[1]] <- c("Admit", "Reject")

data(Arthritis)
art <- xtabs(~ Improved + Treatment, data = Arthritis)
@

<<auxfunctions,echo=FALSE,results=hide>>=
he.test <- pearson.test(haireye, return = TRUE)
cb <- he.test$qdist(c(0.9, 0.99))

gp.shading <- function(x, y) gp.HSVshading(x, y, lty = 1, test = NULL)
shadeHSVchisq <- function(x, y) gp.HSVshading(x, y, lty = 1)
shadeHSVmax <- function(x, y) gp.HSVshading(x, y, interpolate = cb, lty = 1)
shadeHCLmax <- function(x, y) gp.HCLshading(x, y, interpolate = cb, lty = 1)
shadeHSVmax2 <- function(x, y) gp.HSVshading(x, y, interpolate = c(5,10), test = NULL)
shadeHCLchisq <- function(x, y) gp.HCLshading(x, y, lty = 1)

@@

The \proglang{R} side then has to be described in another paper,
probably for the Journal of Statistical Software.\\[0.3cm]
\textbf{Key Words:} Association plots; Conditional inference; 
  Contingency tables; Mosaic plots; Permutation tests.
\end{abstract}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.assocplot(goals, rot = 0, gp = gp.binary)
@
<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaicplot(t(goals), shade = TRUE, main = "")
@
\caption{Classic association and mosaic plot for the Bundesliga data.}
\label{fig:goals-assoc-classic}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
<<haireye-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.mosaicplot(haireye, gp = gp.Friendly)
@
<<haireye-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.mosaicplot(haireye, gp = gp.max)
@
\caption{Mosaic plots for the Hair-Eye-Color data.}
\label{fig:haireye}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
<<goals-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.mosaicplot(goals, gp = gp.Friendly)
@
<<goals-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.mosaicplot(goals, gp = gp.max)
@
\caption{Mosaic plots for the Bundesliga data.}
\label{fig:goals}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<UCB-mosaic1,fig=TRUE,echo=FALSE,height=4,width=8>>=
fm <- loglm(~(Gender + Admit)*Dept, data = UCBAdmissions, fitted = TRUE)
res.fm <- (UCBAdmissions - fitted(fm))/sqrt(fitted(fm))
dim(res.fm) <- dim(UCBAdmissions)
my.gp <- gp.HCLshading(UCBAdmissions, aperm(res.fm))
my.gp$p.value <- summary(fm)$tests[2,3]
my.gp$lty <- c(1,1)
grid.mosaicplot(ucb, gp = my.gp, margin = ~ (Gender + Admit)*Dept, space = c(0.1, 0.01, 0))
@
\caption{UCB admission with mosaic display.}
\label{fig:UCB-mosaic1}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
<<UCB-mosaic2,fig=TRUE,echo=FALSE,height=6,width=8>>=
mygp.max <- function(x,y) gp.max(x, y, level = (1 - c(0.1, 0.01)/6))
tabplot(ucb, panel = function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE, labels = FALSE, gp = mygp.max, space = 0.02, ...))
@
\caption{UCB admission with conditioned mosaic display.}
\label{fig:UCB-mosaic2}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
<<UCB-assoc,fig=TRUE,echo=FALSE,height=8,width=8>>=
tabplot(margin.table(ucb, c(2,1,3)), gp = mygp.max)
@
\caption{UCB admission with conditioned association plot.}
\label{fig:UCB-assoc}
\end{center}
\end{figure}

>>>>

%% Hyperref
\hypersetup{%
  pdftitle = {Residual-based Shadings for Visualizing (Conditional) Independence},
  pdfsubject = {Residual-based Shadings},
  pdfkeywords = {contingency tables, association plots, mosaic plots, HCL colors, 
    HSV colors, conditional inference, permutation tests},
  pdfauthor = {Achim Zeileis, David Meyer, Kurt Hornik},
  %% change colorlinks to false for pretty printing
  colorlinks = {true},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

@@

\begin{center}
\textbf{\LARGE Residual-based Shadings for Visualizing (Conditional) Independence}

@@

data(Arthritis)
art <- xtabs(~ Improved + Treatment, data = Arthritis, subset = Sex == "Female")

data(Employment)
emp <- Employment
emp1 <- co.table(Employment, "LayoffCause")[["Replaced"]]
@

<<auxfunctions,echo=FALSE,results=hide>>=
pmos.Friendly <- function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE,
  labels = FALSE, space = 0.02, gp = gp.Friendly, ...)
pmos.max <- function(x, ...) grid.mosaicplot(x, panel = TRUE, legend = FALSE,
  labels = FALSE, space = 0.02, gp = gp.max, ...)

@@

The \proglang{R} side then has to be described in another paper,
probably for the Journal of Statistical Software.\\[0.3cm]
\textbf{Key Words:} Association plots; Conditional inference; 
  Contingency tables; HCL colors; HSV colors; Mosaic plots; Permutation tests.
\end{abstract}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.assocplot(emp1, gp = gp.binary)
@
<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaicplot(t(emp1), shade = TRUE, main = "")
@
\caption{Classic association and mosaic plot for the ``Replaced'' subset of the Employment data.}
\label{fig:classic}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
<<emp1-Friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.mosaicplot(emp1, gp = gp.Friendly)
@
<<emp1-max,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.mosaicplot(emp1, gp = gp.max)
@
\caption{Mosaic plots for the ``Replaced'' subset of the Employment data.}
\label{fig:emp1}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<UCB-mosaic1,fig=TRUE,echo=FALSE,height=6,width=8>>=
fm <- loglm(~(EmploymentStatus+EmploymentLength)*LayoffCause, data = emp, fitted = TRUE)
my.gp <- gp.HCLshading(emp, residuals(fm, type = "pearson"))
my.gp$p.value <- summary(fm)$tests[2,3]
my.gp$lty <- c(1,1)
grid.mosaicplot(emp, gp = my.gp, margin = ~(EmploymentStatus+EmploymentLength)*LayoffCause,
  space = c(0.1, 0.01, 0), dir = "v")
@
\caption{Employment data with mosaic display.}
\label{fig:emp-mosaic1}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
<<UCB-mosaic2,fig=TRUE,echo=FALSE,height=4,width=8>>=
tabplot(emp, panel = pmos.max)
@
\caption{Employment data with conditioned mosaic display.}
\label{fig:emp-mosaic2}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
<<UCB-assoc,fig=TRUE,echo=FALSE,height=4,width=8>>=
tabplot(emp)
@
\caption{Employment data with conditioned association plot.}
\label{fig:emp-assoc}
\end{center}
\end{figure}
